---
title: "NB Iris Tutorial"
author: "Olivia Sheng"
date: "September 16, 2016"
output: 
  html_document:
    highlight: breezedark
    number_sections: yes
    toc: yes
    fig_width: 10
    fig_height: 5
editor_options: 
  chunk_output_type: console
---
# Data description
This is perhaps the best known database to be found in the pattern recognition literature.The data set contains 3 classes of 50 instances each, where each class refers to a type of iris plant.  One class is linearly separable from the other 2; the latter are NOT linearly separable from each other.

Iris images: https://www.researchgate.net/figure/220371329_fig2_Fig-5-Iris-Setosa-Versicolor-and-Virginica-flowers-The-photos-are-from

Target variable: class of iris plant. 

Number of Instances: 150 (50 in each of three classes)

Number of Attributes: 4 numeric, predictive attributes and the class variable

Attribute Information:
1. sepal length in cm
2. sepal width in cm
3. petal length in cm
4. petal width in cm
5. class: Iris Setosa, Iris Versicolour, Iris Virginica

# Setup and overall data inspection
```{r Setup and overall data inspection}
# Package loading. Install the following packages before running this chunk or knitting this program.

library(e1071)
library(psych)
library(caret)
library(rminer)
library(rmarkdown)
library(tictoc) 
tic()

# R already has an iris data frame. Examine the overall data frame

str(iris)
summary(iris)
```

# NB model building using e1071 package

```{r NB model building}
# e1071 includes a naiveBayes algorithm to build a Naive Bayesian classification model
# e1071 documentation - https://cran.r-project.org/web/packages/e1071/e1071.pdf

# Run ?naiveBayes to find out more information about naiveBayes 


iris_nb_w <- naiveBayes(Species ~ ., data = iris)

## alternatively, the two arguments in naiveBayes are the data frame with predictors only and the target varible colum

iris_nb_w <- naiveBayes(iris[,-5], iris[,5])

# Examine the Means and Sds of the predictors for the iris model

iris_nb_w
```

# Explantory data exploration

```{r Explanatory data exploration}
# We can compare the model info to boxplots that group values of a numeric variable by Species 

boxplot(Petal.Length~Species, data = iris, ylab = "Petal Length")
boxplot(Petal.Width~Species, data = iris, ylab = "Petal Width")
boxplot(Sepal.Length~Species, data = iris, ylab = "Speal Length")
boxplot(Sepal.Width~Species, data = iris, ylab = "Speal Width")

# Separate data by Species class

df_setosa <- subset(iris, Species == "setosa")
summary(df_setosa)
df_versicolor <- subset(iris, Species == "versicolor")
summary(df_versicolor)
df_virginica <- subset(iris, Species == "virginica")
summary(df_virginica)

# Plot predictors' correlations by class

pairs.panels(df_setosa[-5])
pairs.panels(df_versicolor[-5])
pairs.panels(df_virginica[-5])
```

# Generate performance metrics

```{r}
# Generate the predictions for the train set

predicted_species <- predict(iris_nb_w, iris)

mmetric(iris$Species, predicted_species, metric="CONF")

mmetric(iris$Species, predicted_species, metric=c("ACC","TPR","PRECISION","F1"))
```
# Simple hold-out evaluation

```{r Simple hold-out evaluation}
# Use createDataPartition() in caret package to split iris 50%-50% into a train set and a test set

# set seed to a value for createDataPartition(). With the same value and input, 
# the partitions output will be consistent each time the following commands are executed.

set.seed(100)
inTrain <- createDataPartition(iris$Species, p=0.5, list=FALSE)

# inTrain is a list of indices to the rows in the iris data frame

str(inTrain)
# inTrain

# Assign the rows in iris indexed by inTrain to create a train set
# Assign all other rows indexed by -inTrain to create a test set

irisTrain <- iris[inTrain,]
irisTest <- iris[-inTrain,]

# Examine the distributions of Species and other attriutes of train and test sets
# Make sure that they are consistent between train and test sets.

summary(irisTrain)
summary(irisTest)

table(irisTrain$Species)
table(irisTest$Species)

prop.table(table(irisTrain$Species))
prop.table(table(irisTest$Species))

# Use the train set to build a model

iris_nb <- naiveBayes(irisTrain$Species~.,irisTrain)
iris_nb

# Apply the model to the hold-out test set and generate holdout evaluation metrics

predicted_species <- predict(iris_nb, irisTest)

mmetric(irisTest$Species, predicted_species, metric="CONF")
mmetric(irisTest$Species, predicted_species, metric=c("ACC","TPR","PRECISION","F1"))

# End of naiveBayes iris Tutorial
toc()
```
