---
title: "C5.0 Iris Tutorial"
author: "Olivia Sheng"
date: "September 16, 2016"
output:
  html_document:
    theme: readable
    highlight: breezedark
    number_sections: yes
    toc: yes
    fig_width: 10
    fig_height: 5
editor_options: 
  chunk_output_type: console
---
# Data description
This is perhaps the best known database to be found in the pattern recognition literature.The data set contains 3 classes of 50 instances each, where each class refers to a type of iris plant.  One class is linearly separable from the other 2; the latter are NOT linearly separable from each other.

Iris images: https://www.researchgate.net/figure/220371329_fig2_Fig-5-Iris-Setosa-Versicolor-and-Virginica-flowers-The-photos-are-from

Target variable: class of iris plant. 

Number of Instances: 150 (50 in each of three classes)

Number of Attributes: 4 numeric, predictive attributes and the class variable

Attribute Information:
1. sepal length in cm
2. sepal width in cm
3. petal length in cm
4. petal width in cm
5. class: Iris Setosa, Iris Versicolour, Iris Virginica

# Setup and overall data inspection
```{r Setup and overall data inspection}
# Package loading. Install the following packages before running this chunk or knitting this program.

library(C50)
library(caret)
library(rminer)
library(rmarkdown)
library(tictoc) 
tic()

# R already has an iris data frame. Examine the overall data frame

str(iris)
summary(iris)
```

# Build a decision tree using C5.0 in C50 package

```{r Build a decision tree}
# C50 includes functions for building and visualizing a decisioin tree model
# based on Quinlan's C5.0 algorithm
# C50 documentation - https://cran.r-project.org/web/packages/C50/C50.pdf
# Load C50 package. 

# Call C5.0 function to build a classification model.
# The first argument in C5.0: iris$Species~. indicates the target variable is iris$Species
# All other variables are predictors for it
# The second argument indicates that the whole data set - iris is the training set
# Use ?C5.0 to see help info for C5.0 
# Expand iris.C50 in the Environment pane to find out more about the model

iris.c50 <- C5.0(iris$Species~.,iris)

# Enter the model name to find out tree size (i.e. the number of leaves)
# Use plot() to plot the tree. 
# Note: plot() does not plot the tree if the tree is too complex
# Use summary() to find out the tree model expressed in rules, 
# the confusion matrix (or contigency matrix), and the error rate.

iris.c50
plot(iris.c50)
summary(iris.c50)

# An alternative C5.0 call to build the same model.
# The 1st argument in C5.0() specifies the predictors to be used.
# The 2nd argument indicates the target variable.
# Compare the model information to see if the tree is identical to the last one built.

iris.c50 <- C5.0(iris[-5],iris$Species)
iris.c50
plot(iris.c50)
summary(iris.c50)
```
# Post-model-building data exploration
```{r Data visualization}
# Visualize the relationships amongst Petal.Width, Petal.Length and Species 
# to see why it is meaningful to segment the data set by these two predictors

with(iris, plot(Petal.Length,Petal.Width, col=Species, pch = as.numeric(Species), main = "2d scatter plot of iris data", sub = "Petal Width vs Length"))
legend('topright', legend = levels(iris$Species),  col = 1:3, cex = 0.8, pch = 1:3)
```

# Derive classification performance metrics using mmetric() in rminer package.

```{r Generate performance metrics}
# rminer builds, tests and evaluates a variety of classification and regression models.
# rminer documentation - https://cran.r-project.org/web/packages/rminer/rminer.pdf

# predict() applies a model (1st argument) to a testing data set (2nd argument).
# Let's apply it to the whole set that was used to train the model 
# to see the model's performance metrics in training data (i.e., not holdout evaluation)
# Take a look at the structure and summary of predicted.species to understand the output of predict()

predicted.species <- predict(iris.c50, iris)
str(predicted.species)
summary(predicted.species)

# mmetric() generates confusion matrix (3rd argument) based on the true target variable values (1st argument)
# and the predicted target variable values (2nd argument) 

mmetric(iris$Species, predicted.species, metric="CONF")

# Take a look at the applicable classification metrics mmetric() can generate

mmetric(iris$Species, predicted.species, metric="ALL")

# Initially we only focus on a few classification metrics the following mmetric() generates. 
# Remember to use combine - c() when you select more than one metric (or metric group) name

mmetric(iris$Species, predicted.species, metric=c("ACC","TPR","PRECISION","F1"))
```

# Simple hold-out evaluation

```{r Simple hold-out evaluation}
# Only knowing the model's training performance is not sufficient. Let's try a simple hold-out evaluation. 

# Use createDataPartition() in caret package to split iris 50%-50% into a train set and a test set
# caret introduction: https://cran.r-project.org/web/packages/caret/vignettes/caret.pdf
# or use ?caret to find this introduction and other documentation

# set seed to a value for createDataPartition(). With the same value and input, 
# the partitions output will be consistent each time the following commands are executed.

set.seed(100)
inTrain <- createDataPartition(iris$Species, p=0.5, list=FALSE)

# inTrain is a list of indices to the rows in the iris data frame

str(inTrain)
inTrain

# Assign the rows in iris indexed by inTrain to create a train set
# Assign all other rows indexed by -inTrain to create a test set

irisTrain <- iris[inTrain,]
irisTest <- iris[-inTrain,]

# Examine the distributions of Species and other attriutes of train and test sets
# Make sure that they are consistent between train and test sets.

summary(irisTrain)
summary(irisTest)

table(irisTrain$Species)
table(irisTest$Species)

prop.table(table(irisTrain$Species))
prop.table(table(irisTest$Species))

# Use the train set to build a model

iris.c50 <- C5.0(irisTrain$Species~.,irisTrain)
iris.c50
plot(iris.c50)
summary(iris.c50)

# Apply the model to the hold-out test set and generate holdout evaluation metrics

predicted.species <- predict(iris.c50, irisTest)
predicted.species
mmetric(irisTest$Species, predicted.species, metric="CONF")
mmetric(irisTest$Species, predicted.species, metric=c("ACC","TPR","PRECISION","F1"))

# End of C5.0 iris Tutorial

toc()
```
