---
title: "Iris Data Exploration Rmarkdown"
author: "Olivia Sheng"
output:
  html_document:
    theme: readable
    highlight: breezedark
    number_sections: yes
    toc: yes
    fig_width: 25
    fig_height: 15
editor_options: 
  chunk_output_type: console
---
# Data description
This is perhaps the best known database to be found in the pattern recognition literature.The data set contains 3 classes of 50 instances each, where each class refers to a type of iris plant.  One class is linearly separable from the other 2; the latter are NOT linearly separable from each other.

Iris images: https://www.researchgate.net/figure/220371329_fig2_Fig-5-Iris-Setosa-Versicolor-and-Virginica-flowers-The-photos-are-from

Target variable: class of iris plant. 

Number of Instances: 150 (50 in each of three classes)

Number of Attributes: 4 numeric, predictive attributes and the class variable

Attribute Information:
1. sepal length in cm
2. sepal width in cm
3. petal length in cm
4. petal width in cm
5. class: Iris Setosa, Iris Versicolour, Iris Virginica

```{r Setup and overall data inspection}
### R already has an iris data frame
library(tictoc) 
tic()

### Remeber to get help from the Help Pane using "?CommandName"

?iris

### Examine the overall data frame

# str() shows the number of observations, and the number, names, types and some values of columns

str(iris)


## You can retrieve and save the number of rows and number of coloumns of a data frame

nrow(iris)
row <- nrow(iris)
row

col <- ncol(iris)
col

## Show the head and tail rows of a data frame

head(iris)

iris[1:6,]

head(iris, n=1)

iris[1,]

head(iris, n=10)

tail(iris, n = 10)

?head

# summary() shows the mean and the five-number statistics indicating the spread of each column's values

summary(iris)
```
# Explore single numeric variables
```{r Explore single numeric variables}
### understanding a single variable: numerical variables

## Show summary of one or more columns

summary(iris$Sepal.Length)

summary(iris[c("Sepal.Length", "Sepal.Width")])

## obtain the mean, median, max, min and range of a numeric variable

mean(iris$Sepal.Length)
median(iris$Sepal.Length)


range(iris$Sepal.Length)

max.Sepal.L <- max(iris$Sepal.Length)
min.Sepal.L <- min(iris$Sepal.Length)

min.Sepal.L
max.Sepal.L

# calculate the difference of the range

diff.Sepal.L <- max.Sepal.L - min.Sepal.L

diff.Sepal.L

diff(range(iris$Sepal.Length))

# min-max normalization of first observation's Sepal.Length to a value between zero zna 1

iris$Sepal.Length[1]

(iris$Sepal.Length[1] - min.Sepal.L)/diff.Sepal.L

# use quantile to calculate the five-number summary

quantile(iris$Sepal.Length)

# the 1st percentile and the 99th percentile

quantile(iris$Sepal.Length, probs = c(0.01, 0.99))

# quintiles and deciles 

quantile(iris$Sepal.Length, seq(from = 0, to = 1, by = 0.20))

quantile(iris$Sepal.Length, seq(from = 0, to = 1, by = 0.10))

# min-max normalization of Sepal.Length
iris$Sepal.Length.n <- (iris$Sepal.Length - min.Sepal.L)/diff.Sepal.L

# use quantile to calculate the five-number summary for normalized Sepal.Length

quantile(iris$Sepal.Length.n)

summary(iris$Sepal.Length.n)

# the 1st percentile and the 99th percentile of normalized Sepal.Length

quantile(iris$Sepal.Length.n, probs = c(0.01, 0.99))

# quintiles and deciles of normalized Sepal.Length

quantile(iris$Sepal.Length.n, seq(from = 0, to = 1, by = 0.20))

quantile(iris$Sepal.Length.n, seq(from = 0, to = 1, by = 0.10))


# boxplot of a numeric variable

boxplot(iris$Sepal.Length, main = "Boxplot of Sepal Length in the iris data set",
        ylab = "Sepal Length")
?IQR

### IQR = 3rdQaurtile - 1stQuartile

### maxline = 3rdQuartile + 1.5*IQR

### minline = 1stQuartile - 1.5*IQR

### one definition of outliers : value > maxline and value < minline are drawn as circles and dots 

# scatter plot of a numeric variable. It becomes too busy for a large number of observations

plot(iris$Sepal.Length)

# histograms of a numeric variable

hist(iris$Sepal.Length, main = "Histogram of Sepal Length in the iris data set",
     xlab = "Sepal Length")

### For a right skewed distribution, the mean is typically greater than the median

hist(iris$Sepal.Width, main = "Histogram of Sepal Width in the iris data set",
     xlab = "Sepal Width")

hist(iris$Petal.Length, main = "Histogram of Petal Length in the iris data set",
     xlab = "Petal Length")

hist(iris$Petal.Width, main = "Histogram of Petal Width in the iris data set",
     xlab = "Petal Width")

### Search for and understand the meaning of skewed, bimodal and multimodal continuous distributions

## variance and standard deviation of a numeric value

var(iris$Sepal.Length)

sd(iris$Sepal.Length)

var(iris$Sepal.Width)

sd(iris$Sepal.Width)
```
# Explore factor variables
```{r Explore factor variables}

# A factor's distinct values

is.factor(iris$Species)

summary(iris$Species)

nlevels(iris$Species)

str(iris$Species)

is.factor(as.character(iris$Species))

summary(as.character(iris$Species))

# How many of observations contain a factor value?

?table

table(iris$Species)

plot(iris$Species)

plot(iris$Species, main = "Plot of Species in the iris data set",
     xlab = "Species")

# Why doesn't the following command work?

# barplot(iris$Species,main = "Barplot of Species in the iris data set",
#       xlab = "Species")

Species.table <- table(iris$Species)

Species.table

barplot(Species.table,main = "Barplot of Species in the iris data set",
        xlab = "Species")

# compute table proportions

?prop.table

# Run prop.table(iris$Species) to see the error in this command

prop.table(Species.table)

Species.prop <- prop.table(table(iris$Species))

Species.prop

## round decimals 

round(Species.prop, digits = 2)

# another way 

options(digits = 2)
prop.table(Species.table)

###Remeber to get help from Help Pane using "?"

?round
?options
```

# Understand relatioinships of multiple variables
```{r Understand relatioinships of multiple variables}
# cor,  boxplot, 2D scatter plot - plot, 3D scatter plot

# scatter plot: two numeric variables

plot(iris[,1:2])

### Generate correlation coefficients of two numeric variables in a 2x2 matrix
# cor(X,Y) lies between -1 and 1. zero means no correlation. 1 or -1 indicates full correlation
# positive value means positive correlation and negative values mean negative relationships
# Examine the components in the formulation for correlation coefficients
# cor(X,Y) = cov(X,Y)/(sd(X)*sd(Y)) 
# cov(X,Y) = E[X-E(X)]*E[Y-E(Y)]

cov(iris[1:2])
var(iris[1:2])
var(iris[,1])
sd(iris[,1])
var(iris[2])
sd(iris[,2])
cor(iris[c("Sepal.Length", "Sepal.Width")])
cor(iris[1:2])
cor(iris[,1:2])

?cor

## Generate the correlation matrix of all numeric variables

cor(iris[1:4])

# Think about the reason why cor(iris) doesn't run

# Generate 2D scatter plots and correlation coefficients

# Load the psych package. Install it first when necessary.

pairs(iris[1:4])

pairs(iris)

library(psych)

pairs.panels(iris[1:4])

pairs.panels(iris)

##### Examine relationships between numeric variables and factors

### boxplot groups values of a numeric variable based on the values of a factor

boxplot(Petal.Length~Species, data = iris)
boxplot(Petal.Width~Species, data = iris)
boxplot(Sepal.Length~Species, data = iris)
boxplot(Sepal.Width~Species, data = iris)

### The aggregate function

# We can use the aggregate command to aggregate a numeric feature by a categorical one.

# The aggregate function has three parameters

# 1. The numeric value, e.g. sales, to be aggregated to find out, e.g., total of sales,
#   average of sales, number of sales (i.e. orders).

# 2. The set of categories, product_category and sales_region, on which you wish
#   to aggregate

# 3.The aggregation function (e.g., sum, mean, length) that you wish to use

aggregate(Petal.Length~Species, summary, data = iris)

aggregate(Petal.Width~Species, summary, data = iris)


# scatter plot of numeric values and factor values

plot(iris$Species)

plot(iris$Sepal.Length)

plot(iris$Sepal.Length,iris$Species)

plot(iris$Sepal.Length,iris$Sepal.Width, col = iris$Species, pch = as.numeric((iris$Species)))

with(iris, plot(Sepal.Length,Sepal.Width, col = Species, pch = as.numeric(Species)))

with(iris, plot(Petal.Length,Petal.Width, col = Species, pch = as.numeric(Species), 
                main = "2d scatter plot of iris data", sub = "Petal Width vs Length"))

palette()

legend('topright', legend = levels(iris$Species),  col = 1:3, cex = 0.8, pch = 1:3)


# Generate 3D scatterplot

library(scatterplot3d)

scatterplot3d(iris$Petal.Length,iris$Petal.Width,iris$Sepal.Length, highlight.3d = TRUE, pch = as.numeric(iris$Species), main = "3D scatter plot of iris data")


### Another parallel plot and parallel coordinate plot to show relationships between numeric variables and factors

### load the required libraries - lattice and MASS
### Install these package first or use require(lattice) and require(MASS) to install and load them


library(lattice)

parallelplot(~iris[1:4] | Species, data = iris, var.label = TRUE)


# Generate parallel coordinate plot

library(MASS) 

parcoord(iris[1:4], col = iris$Species, var.label = TRUE)

palette()

legend('topright', legend = levels(iris$Species),  col = 1:3, cex = 0.8, pch = 1:3)
```

# Import and Export CSV files
```{r Import and Export CSV files}
# setwd specifies working directory where input data and output files are located. For example

cloud_wd <- getwd()
setwd(cloud_wd)

#Export to a csv file

write.csv(iris,file = "irisfile.csv")

?write.csv

##Export to a file of xls extension by using WriteXLS package
# install the package "WriteXLS"
# install.packages("WriteXLS") must be run before rendering an RMarkdown program.
# Load the package
library("WriteXLS")
?WriteXLS
WriteXLS(iris, ExcelFileName = "irisWriteXLS.xls")


# Read irisfile.csv into a data frame. 

iris2 <- read.csv("irisfile.csv", stringsAsFactors = FALSE)

str(iris2)

# What are the differences between iris2 different from iris?

# omit saving row.names or row numbers and change Species to a factor

write.csv(iris,file = "irisfile.csv", row.names = FALSE)

iris2 <- read.csv("irisfile.csv", stringsAsFactors = FALSE)

str(iris2)

iris2$Species < factor(iris2$Species)

str(iris2)


# It is a good idea to specify the full path of the file.

iris2 <- read.csv("irisfile.csv")

str(iris2)

# Store the file name 


inputfile <- "irisfile.csv"

iris2 <- read.csv(file = inputfile)

str(iris2)

# Find help document by using "?"

?read.csv

# In a rmarkdown program, you cannot browse to select the file to read

# iris2 <- read.csv(file.choose())

# str(iris2)

### End of Iris Data Exploration Tutorial
toc()

```

